---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this. You could also use GitHub Actions to re-render `README.Rmd` every time you push. An example workflow can be found here: <https://github.com/r-lib/actions/tree/master/examples>. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# canaper

<!-- badges: start -->
[![DOI](https://zenodo.org/badge/359280907.svg)](https://zenodo.org/badge/latestdoi/359280907)
[![R-CMD-check](https://github.com/joelnitta/canaper/workflows/R-CMD-check/badge.svg)](https://github.com/joelnitta/canaper/actions)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Codecov test coverage](https://codecov.io/gh/joelnitta/canaper/branch/main/graph/badge.svg)](https://codecov.io/gh/joelnitta/canaper?branch=main)
<!-- badges: end -->

```{r srr-tags-intro, eval = FALSE, echo = FALSE}
#' @srrstats {G1.1} documents novelty
```

The goal of canaper is to enable [categorical analysis of neo- and paleo-endemism (CANAPE)](https://doi.org/10.1038/ncomms5473) in **R**. This is the first implementation in **R** of CANAPE, which was previously only available in [Biodiverse](http://shawnlaffan.github.io/biodiverse/).

## Important note

**This package is in early development.** There may be major, breaking changes to functionality in the near future. If you use this package, I highly recommend using a package manager like [renv](https://rstudio.github.io/renv/articles/renv.html) so that later updates won't break your code.

## Installation

You can install canaper from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("joelnitta/canaper")
```

## Example usage

These examples use the dataset from [Phylocom](http://phylodiversity.net/phylocom/). 
The dataset includes a community (site x species) matrix and a phylogenetic tree.

```{r example-data}
library(canaper)

data(phylocom)

# Example community matrix including 4 "clumped" communities,
# one "even" community, and one "random" community
phylocom$comm

# Example phylogeny
phylocom$phy
```

The main "workhorse" function of `canaper` is `cpr_rand_test()`, which conducts a randomization test to determine if observed values of phylogenetic diversity (PD) and phylogenetic endemism (PE) are significantly different from random. It also calculates the same values on an alternative phylogeny where all branch lengths have been set equal (alternative PD, alternative PE) as well as the ratio of the original value to the alternative value (relative PD, relative PE).

```{r rand-test-example}
set.seed(071421)
rand_test_results <- cpr_rand_test(
  phylocom$comm, phylocom$phy,
  null_model = "swap"
)
```

`cpr_rand_test` produces **a lot** of columns (nine per metric), so let's just look at a subset of them:

```{r rand-test-res}
rand_test_results[, 1:9]
```

This is a summary of the columns:

- `*_obs`: Observed value
- `*_obs_c_lower`: Count of times observed value was lower than random values
- `*_obs_c_upper`: Count of times observed value was higher than random values
- `*_obs_p_lower`: Percentage of times observed value was lower than random values
- `*_obs_p_upper`: Percentage of times observed value was higher than random values
- `*_obs_q`: Count of the non-NA random values used for comparison
- `*_obs_z`: Standard effect size (z-score)
- `*_rand_mean`: Mean of the random values
- `*_rand_sd`: Standard deviation of the random values

The next step in CANAPE is to classify endemism types according to the significance of PE, alternative PE, and relative PE. This adds a column called `endem_type`.

```{r categorize-endem}
canape_results <- cpr_classify_endem(rand_test_results)

canape_results[, "endem_type", drop = FALSE]
```

This data set is very small, so it doesn't include all possible endemism types. In total, they include:

- `paleo`: paleoendemic
- `neo`: neoendemic
- `not significant` (what it says)
- `mixed`: mixture of both paleo and neo
- `super`: mixed and highly significant (*p* < 0.01)

For a more complete example, please [see the vignette](https://joelnitta.github.io/canaper/articles/canape.html)

## Comparsion with other software

Several other R packages are available to calculate diversity metrics for ecological communities. The non-exhaustive summary below focuses on alpha diversity metrics in comparison with `canaper`, and is **not** a comprehensive description of each package.

- [PhyloMeasures](https://doi.org/10.1111/ecog.01814): Calculates phylogenetic community diversity metrics including MPD, MNTD, PD, phylosor, and unifrac. Null models for matrix randomization include `uniform`, `frequency.by.richness`, and `sequential`.
- [phyloregion](https://github.com/darunabas/phyloregion): Calculates PD but not MPD or MNTD. Implements sparse matrix encoding to increase computing speed, which is used by `canaper`. Null models for matrix randomization include `tipshuffle`, `rowwise`, and `colwise`. Also performs regionalization based on taxonomic or phylogenetic beta diversity.
- [picante](https://github.com/skembel/picante): Calculates MPD, MNTD, PD, etc. Null models for community matrix randomization include `frequency`, `richness`, `independentswap`, and `trialswap`.
- [vegan](https://github.com/vegandevs/vegan): Performs a large range of mostly non-phylogenetic diversity analyses. Includes the largest selection of null models (> 20), according to data type (binary vs. quantitative). `canaper` uses `vegan` to randomize community matrices.
- [biodiverse](http://shawnlaffan.github.io/biodiverse/): Not an R package, but software written in perl with a GUI. Performs all of the calculations needed for CANAPE, and many other metrics (> 300). Includes `rand_structured` null model as well as spatially structured null models. None of these null models are currently available in any R packages AFAIK, except for `independentswap`.

## Other information

Poster at [Botany 2021](https://2021.botanyconference.org/)

- [PDF](https://github.com/joelnitta/botany_poster_2021/raw/main/canaper.pdf)
- [Video](https://www.youtube.com/watch?v=LXn_ziEli_Q)

## Citing this package

If you use this package, please cite it! Here is an example:

```
Nitta JH, Laffan SW, Mishler BD, Iwasaki W. (2021) canaper: Categorical analysis of neo- and paleo-endemism in R. doi: 10.5281/zenodo.5094032
```

The example DOI above is for the overall package. 

Here is the latest DOI, which you should use if you are using the latest version of the package:

[![DOI](https://zenodo.org/badge/359280907.svg)](https://zenodo.org/badge/latestdoi/359280907)

You can find DOIs for older versions by viewing the "Releases" menu on the right.

## Papers citing `canaper`

- Nitta *et al.* "Spatial phylogenetics of Japanese ferns: Patterns, processes, and implications for conservation". bioRxiv https://doi.org/10.1101/2021.08.26.457744

## Contributing and code of conduct

Contributions to `canaper` are welcome! For more information, please see [CONTRIBUTING.md](.github/CONTRIBUTING.md)

Please note that this package is released with a [Contributor Code of Conduct](https://ropensci.org/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## Licenses

- Code: [MIT](LICENSE.md)
- Example datasets
  - `acacia`, `biod_example`: [GNU General Public License v3.0](https://github.com/joelnitta/canaper/blob/main/data-raw/LICENSE-gpl.txt)
  - `phylocom`: [BSD-3-Clause](https://github.com/joelnitta/canaper/blob/main/data-raw/LICENSE-bsd3.txt)

## References

Mishler, B., Knerr, N., González-Orozco, C. *et al*. Phylogenetic measures of biodiversity and neo- and paleo-endemism in Australian *Acacia*. *Nat Commun* 5, 4473 (2014). https://doi.org/10.1038/ncomms5473
