---
title: "CANAPE example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CANAPE example}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette replicate the analysis of Mishler 2014, where CANAPE was originally defined. Let's load `canaper` and some other packages we will use for this vignette.

```{r setup}
library(canaper)
library(ape)
```

The data are included in the `canaper` package. Let's load the data into memory.

```{r load-data}
data(acacia)
```

The `acacia` dataset is a list including a phylogeny of *Acacia* species in Australia and a community dataframe with species as columns and rows as sites. The row names correspond to the centroids of 50 x 50 km grid cells covering Australia.

```{r inspect-data}
acacia$phy

# The community matrix is too large to print out in its entirety, so
# we will just take a look at the first 8 rows and columns
dim(acacia$comm)
acacia$comm[1:8,1:8]
```

